### 2022-02-16 README.Rmd
begin with the main argument

Preregistration is becoming increasingly popular in the psychological sciences.
At the same time, many researchers struggle to make precise predictions and feel overwhelmed by the need to prespecify every detail of their analysis plan.
Should they surrender and discard the idea of preregistration altogether? Not at all.
We argue for the benefits of preregistration beyond strictly confirmatory studies.
Grounded in Bayesian philosophy of science, we define a formal objective for preregistration that neither declares changes to a preregistration to be sinful nor punishes rigor.
Our argument rests on the relevance of “theoretical risk”, a generalization of type-I error rate, for judging theories.
Our proposal is that preregistration should aim to reduce the uncertainty in judging the theoretical risk instead of limiting the type-I error rate.
We argue that this reformulation extends the utility of preregistration and is closer aligned with researchers intuition.
If we do not know what a researcher did, we have great uncertainty about the theoretical risk they took.
This uncertainty about the theoretical risk does us a disservice by decreasing the evidence boost.

A successful prediction should lend more credibility to a theory than a postdiction, all else being held equal.
Much has been theorized on this important distinction between exploration and confirmation, discovery and justification, hypothesis generating and hypothesis testing and so forth (see XXXX).
To honor this distinction, many psychological studies are preregistered to produce results that are considered confirmatory.
This appreciation of confirmatory results falls neatly in line with Bayesian reasoning (XXX).
We leverage this reasoning to outline two paradigms that rationalize the use of preregistration.

>>>>

Preregistration is becoming increasingly popular in the psychological sciences.
At the same time, many researchers struggle to make precise predictions and feel overwhelmed by the need to prespecify every detail of their analysis plan.
Should they surrender and discard the idea of preregistration altogether? Not at all.
We argue for the benefits of preregistration beyond strictly confirmatory studies.
Grounded in Bayesian philosophy of science, we define a formal objective for preregistration that neither declares changes to a preregistration to be sinful nor punishes rigor.
Our argument rests on the relevance of “theoretical risk”, a generalization of type-I error rate, for judging the evidential support for theories.
Our proposal is that preregistration should aim to reduce the uncertainty in judging the theoretical risk instead of directly limiting the type-I error rate.
We argue that this reformulation extends the utility of preregistration and is closer aligned with researchers intuition.
If we do not know what a researcher did, we have great uncertainty about the theoretical risk they took.
This uncertainty severely limits the utility of the researchers' evidence.

A successful prediction should lend more credibility to a theory than a postdiction, all else equal.
Much has been theorized on this important distinction between exploration and confirmation, discovery and justification, hypothesis generating and hypothesis testing and so forth (see XXXX).
To honor this distinction, many psychological studies are preregistered to produce results that are considered confirmatory.
Rigorous application of preregistration prevents researcher from reporting a simple confirmatory story for a set of results that were produced by an arduous process of try and error.
The scientific community, widely agrees that actively concealing such "trying for results" is misconduct.
But what exactly is such concealment cheating the scientific community of?
We believe that a rigorous answer to this question enables us to clear some common misunderstandings that lead researchers to deem preregistration as unfit for their purposes.
These misunderstandings are interesting from a theoretical perspective but also lead to decreased efficiency of the scientific process if researcher  misguidedly decide that preregistration is not applicable for their intended study.

One reason why researchers are hesitant to preregister is that they feel too limited by preregistration.
Even researchers who have been persuaded by its advantages sometimes apply strategies to skirt the self imposed restrictions.
One strategy is to write a loose preregistration to begin with, another is to simply deviate from the preregistration afterward.
To complicate matters further, both strategies may be used with compelling scientific reasoning or with the self-serving intent of generating good looking results no matter the nature of the phenomenon under study.
To summarize, the conundrum is that current conceptualization of preregistration deem a study "properly" preregistrated even if it contains a recipe for p-hacking before data collection but "improper" if the researcher did technically deviate afterwards.
This, of course, goes against common researchers intuition about what is considered exploratory and confirmatory.
We point these issues out as symptoms of a misunderstanding.#change
We do want not to discredit preregistration as a useful practice for advancing science, but challenge the common view that preregistration somehow cleanly separates confirmatory from exploratory.

Deciding if a study is confirmatory or exploratory is a judgment made by the researcher.
The aforementioned issues arise by delegating this judgment to a simple decision rule, that is, if something is preregistered it is confirmatory and if it is not preregistered it is exploratory.
Preregistering an analysis that is inherently exploratory (like testing dozens of relations) does not make it confirmatory nor will a careful confirmatory study become exploratory if researcher deviate from the preregistration in minor details.
To uphold the simple decision rule, it is routinely suggested to only preregister when a study is expected to have a low type I error rate by design and will run/be analysed without changes no matter what.
Under these conditions the decision rule holds, but such restricted use makes preregistration a niche solution not able to match the greater problem of replicability in psychology and elsewhere.

We show that this simple decision rule is a special case of a more general conceptualization under Bayesian reasoning.
To that end, we first introduce some basic tools of Bayesian philosophy of science and map the exploration/confirmation distinction onto a dimensional quantity we call "theorethical risk", that is inversely related to type I error rate.
We than propose two interpretations of how theorethical risk is influenced by preregistration.
The first interpretation corresponds to the traditional application of preregistration to research paradigms that focus on confirmation by maximizing the theoretical risk or equivalently limiting type I error.
The second interpretation, is our main contribution, and shows broad applicability of preregistration for both exploratory and confirmatory studies that are implemented as preregistered or have undergone changes.
Under this interpretation theoretical risk is not necessarily directly maximized by preregistration but the uncertainty in judging the theoretical risk is minimized.
This changed viewpoint is crucial for the following reason.
Directly maximizing theoretical risk does, as the name implies, increase the risk that a study does fail to deliver results in favor of the theory.
Minimizing the uncertainty around the theoretical risk, as we will later show, shares the beneficial properties but without increasing likelihood of a "failed" study.
Let us note, that a study that is not able to find evidence that a researcher is hoping for, is still immensely valuable for the scientific process.
Under the current incentive structure, however, many researchers have reason to avoid such outcome and few, if any, have the privalige to operate outside of these incentives.
We therefore propose an interpretation of preregistration that increases the expected epistemic value of studies without necessarily increasing the theoretical risk.

## The epistemic value of studies from a bayesian perspective

Let us start by defining what we mean by expected epistemic value.
If researchers plan to conduct a study they usually hope it will change their assessment of the verisimilitude (truthlikeness) of some theory.
In other words, they hope to learn something from conducting the study.
The amount of knowledge researchers gain from a particular study in relation to a specific theory is what we call epistemic value.
While researcher can not know what exactly they will learn from a study, they can form an expectation.
This expectation is what we call expected epistemic value.
We assume three things about this process of estimation and how it relates to choosing a study to conduct.

1. Researchers judge evidence for or against a hypothesis rationally.
2. They expect other researchers to apply the same rational process.
2. All else equal, researchers try to increase the expected epistemic value for other researchers.

The first assumption leads to our adoption of a Bayesian framework.
Our rational is as follows.
Researchers that decide to conduct a study are akin to choosing a study to bet on.
They have to "place the bet" by conducting the study therefore invest resources and stand to gain epistemic value with some probability.
This conceptualization of choosing a study to conduct as a betting problem, lets us apply a "Dutch Book" argument.
This argument, states that any better must follow the axioms of probability if they want to avoid being "irrational", i.e., accept bets with sure losses.
Fully developing a Dutch book argument for this problem requires a careful consideration of what kind of studies to include as possible bets, defining a conversion rate from the stakes to the reward, and modeling what liberties researchers have in choosing a study.
Without deliberating these concepts further, we find it persuasive that researchers should not violate the axioms of probability if they have some expectation about what they stand to gain with some likelihood from conducting a study.
The axioms of probability are all that is required to derive the famous Bayes formula, on which we will heavily rely for our arguments.
Please note that our decision to adopt a Bayesian philosophy of science does not imply anything about the statistical methods researcher use.
In fact, this conceptualization is purposefully reductionistic to be compatible with a wide range of philosophies of science and statistical methods researchers might subscribe to.
Our conclusions can be applied to bayesian and frequentist methods alike.
Of course, frequentist methods are traditionally viewed from a poperian philosophy of science, but our conclusions are compatible with it and notable later advancements like the error statistical view.
However, those philosophies of science require a much richer set of assumptions that are not quite necessary for the points we want to make.

## Relating Riskiness to epistemic value

The first point is that risky predictions create persuasive evidence, if they turn out to be correct.
This point is important because we attribute much of the appeal of preregistration to this fact.

Let us make some simplifying assumptions and define notation.
We restrict ourselfs to consider evidence of a binary nature (either exists or does not), since continues evidence would led to some quite involved derivations.
We denote the probability of a theory before observing evidence as $P(H)$ and its complement as $P(¬H) = 1 - P(H)$.
The probability of observing evidence under some hypothesis is $P(E|H)$.
The total probability of evidence, $P(E)$, is given by $P(H)P(E|H) + P(¬H)P(E|¬H)$.
From these probabilities we can calculate the probability of the hypothesis after observing the evidence via:

$$
P(H|E) = \frac{P(H)P(E|H)}{P(H)P(E|H) + P(¬H)P(E|¬H)}
$$

From the three probabilities involved in the calculation $P(E|¬H)$ is the one our argument rests on.

The other two assumptions are later necessary to connect individual decisions that effect epistemic value to the research community as a whole.

Observe, that since $P(E|¬H)$

Please note, that we adopt a Bayesian rational for the meta scientific process of preregistration but that this does not imply any ties to the methods a researcher uses.

However, we aim to show that preregistration is indispensable for adequately judging a much broader range of studies.
To that end, we first rationalize the appreciation for conformation using Bayesian reasoning.
Equipped with some of the basic tools of Bayesian Philosophy of Science, we can move beyond a simple dichotomy of exploration and confirmation.

To that end we show that the appreciation for the distinction between exploration and confirmatory neatly falls in line with Bayesian reasoning connected through a quantity we call *theorethical risk*.
Than we proceed to show that preregistration impacts the theorethical risk via two pathways.
Only considering the one path rationalizes recommendation to employ preregistration only for confirmatory studies, taking both paths together into consideration leads to a much wider applicability of preregistration.

But what exactly have you been cheated of?
This appreciation of confirmatory results falls neatly in line with Bayesian reasoning (XXX).
<!-- We leverage this reasoning to outline two paradigms that rationalize the use of preregistration. -->
In order to apply Bayesian reasoning we have to cast conformation and exploration in terms of probability.
Consider the hypothesis that people who regularly engage in fitness activities are healthier.
A confirmatory story might operationalize regularly as weekly, fitness activity as light jogging and healthy as blood cholesterol level.
A more exploratory study might consider several definitions of "regularly", "fitness activities", and "healthy", e.g. "monthly", "weekly", "daily", or "marathon", "light jogging", "intense cleaning", or "cholesterol", "self rated health", "ability to play tag".
Both studies have some probability to turn up with positive results even in a world where fitness activities do not lead to healthier people, because there are other plausible explanations for the results.
However, for the second study the number of explanations that could lead to evidence in favor, aside from the theory is much greater.
Suppose a researcher actually conducted the second study but is reporting it as if they conducted the first study.
You would feel cheated.
But what exactly have you been cheated of?
In fact, you have been cheated out of the information about all the possibilities that could have brought about this evidence.
We summarize these possibilities as a probability.
Therefore, we argue that confirmatory studies have a low probability of observing evidence in favor of their theory if we assume the theory to be false.
Exploratory studies on the other hand are characterized by a higher probability that they find evidence, even when their hypothesis turns out to be wrong.

We connect the question of exploration vs confirmation to Bayesian reasoning via a quantity we call theoretical risk (we borrow the term from Meehl).

Theoretical risk is the inverse of the probability that one observes evidence in favor of a theory when we assume that the theory does not hold.
We assume that confirmatory studies take a high theoretical risk, compared to exploratory studies which take a low theoretical risk.
The tighter your definition of "evidence in favor" is, the more the probability of observing the evidence is tied to your theory holding.
